Sufficient Conditions for Reachability in Automata Networks with Priorities

### 2015-01-24 parts/sa.tex
sa++

% vi:spell spelllang=en:
\section{Static Analysis}\label{sec:sa}

\todo{Replacer notre méthode par rapport aux méthodes d'interprétation abstraite en général}

The aim of this section is to define the problem of reachability in an AAN,
and propose an under-approximation allowing to efficiently solve it.
The static analysis presented here is inspired from~\cite{PMR12-MSCS}.
We consider in this section an AAN $\PH = (\PHs; \PHl; \PHa)$.

\medskip

\modMF{
The main idea behind the static analysis presented in this section
is to abstract the dynamics of an AAN model
by a dynamics that is more general and easier to analyse.
For this, we focus on the notion of objective:
an objective is a couple of local states (\eg $\PHobj{a_i}{a_j}$) whose reachability
is required to ensure the global reachability property considered;
in other words, it is required that a set of actions hitting $a$ exists so that,
starting from a state containing $a_i$,
it is possible to play these actions (possibly intertwined with other actions)
and reach a state containing $a_j$.
Given the particular form of the actions in AANs,
the solving of such an objective raises new objectives in the general case.
%because each action can be triggered by (at most) one local state from a different automaton.
Indeed, each of these local states also have to be reached before the required actions are played.
However, the order of the considered actions is abstracted,
and so is the order of the related objectives,
thus resulting in an over-approximation of the requirements.
}

\subsection{Preliminary Definitions}
\label{ssec:sa-def}

The reachability of a local state $a_j$ of a given automaton $a$,
starting from another local state $a_i$,
is called an objective and is denoted $\PHobjp{a}{i}{j}$ (\pref{def:obj}).
\begin{definition}[Objective ($\Obj$) \& Objective Sequence ($\OS$)]
\label{def:obj}
  If $a \in \components$, the reachability of a local state $a_j$ from a local state $a_i$ is called an \emph{objective}, noted $\PHobj{a_i}{a_j}$.
  The set of all objectives is called $\Obj \DEF \{ \PHobj{a_i}{a_j} \mid a \in \components \wedge (a_i, a_j) \in \PHl_a \times \PHl_a \}$.
  For an objective $P = \PHobj{a_i}{a_j} \in \Obj$, we define: $\PHsort(P) \DEF
  a$, $\PHtarget(P)\DEF a_i$, $\PHbounce(P)\DEF a_j$,
  and $P$ is said \emph{trivial} if $a_i = a_j$.

@@

A context (\pref{def:context}) extends the notion of state to a set of possible initial states:
\modMF{
to each automaton in the model, a context maps a set of local states in this automaton.
}
We also extend the override operator to contexts (\pref{def:ctxcap}).
\begin{definition}[Context ($\Ctx$)]
\label{def:context}
  A \emph{context} $\ctx$ associates to each automaton in $\PHs$ a non-empty subset of its local states:
  $\forall a \in \PHs, \PHget{\ctx}{a} \subseteq \PHl_a \wedge \PHget{\ctx}{a} \neq \emptyset$.
  $\Ctx$ is the set of all contexts.
\end{definition}
%
\begin{definition}[$\Cap: \Ctx \times \powerset(\PHproc) \rightarrow \Ctx$]
\label{def:ctxcap}
  For any $\ctx\in\Ctx$ and set of local states $ps \in \powerset(\PHproc)$,
  the override of $\ctx$ by $ps$ is noted $\ctx \Cap ps$ and is defined by:
  \[ \forall a \in \PHs, \PHget{(\ctx \Cap ps)}{a} \DEF
  \begin{cases}
    \{ p \in ps \mid \PHsort(p)=a \} & \text{if } \exists p \in ps, \PHsort(p)=a,\\
    \PHget{\ctx}{a} & \text{else.}
  \end{cases}
  \]
\end{definition}
\noindent
For a given context $\ctx$, we note $a_i \in \ctx$ if and only if $a_i \in \PHget{\ctx}{a}$,
and for all $ps \in \powerset(\PHproc)$ or $ps \in \PHl$, $ps \subseteq \ctx \EQDEF \forall a_i \in ps, a_i \in \ctx$.
A sequence of actions $\delta$ is \emph{playable} in a context $\ctx$ if and only if 
$\exists s \subseteq \ctx, \delta \in \Sce(s)$.
We denote then: $\delta \in \Sce(\ctx)$,
and the play of $\delta$ in $\ctx$ is $\ctx \PHplay \delta = \ctx \Cap \ceil(\delta)$,
where $\ceil(\delta)$ is the set containing the last local state in the sequence $\delta$ (hitter or bounce) of every automaton mentioned in $\delta$.

Finally, a bounce sequence on a automaton $a$ (\pref{def:bs}) is a sequence of actions hitting $a$
in which the bounce of each action equals the hitter of the following action.
Bounce sequences are used to find local solutions to a given objective;
\modMF{
indeed, the bounce sequences solving an objective $\PHobj{a_i}{a_j}$
only depend on the bounce part of the actions hitting the automaton $a$
and not on the hitters.
}%
Then, a bounce sequence on $a$ can be abstracted into
its combinations of hitters outside automaton $a$ (\pref{def:aBS});
% the minimal sets of its sets of hitters that are not in automaton $a$ (\pref{def:aBS}).
This abstraction is later used to propagate an objective on a given automaton
by creating new objectives on other automata,
by considering the sets of hitters required by the actions involved.

\begin{definition}[Bounce sequence ($\BS$)]
\label{def:bs}
  A \emph{bounce sequence} $\zeta$ is a sequence of actions so that $\forall n \in \indexes{\zeta}, n < |\zeta|, \PHbounce(\zeta_{n}) = \PHtarget(\zeta_{n+1})$.
  $\BS$ denotes the set of all bounce sequences, and
  $\BS(P)$ denotes the set of bounce sequences \emph{solving} an objective $P$:
  \[
    \BS(\PHobj{a_i}{a_j}) \DEF \{ \zeta \in \BS \mid \PHtarget(\zeta_1)= a_i \wedge \PHbounce(\zeta_{|\zeta|}) = a_j \} \enspace.
  \]
  Furthermore, $\BS(\obj{a_i}{a_j}) = \emptyset$ if there is no way to reach $a_j$ from $a_i$
  and $\emptyseq \in \BS(\obj{a_i}{a_i})$
  \modMF{
  for any trivial objective.
  }
\end{definition}

In the following, we denote: $\sSol = \powerset(\PHproc)$
\moda{%
and $\Sol = \powerset(\sSol)$.
}%
\modMF{%
The elements of $\sSol$ are sets of local states,
and will later represent sets of hitters of a given action
that are required to be active simultaneously in order to play this related action.
The elements in $\Sol$ are sets of sets of hitters,
and represent the requirements to play a whole bounce sequence.
}%

@@

\subsection{Under-approximation}
\label{ssec:ua}

We denote by $\concr(\w)$ (\pref{def:concr}) the set of scenarios concretising an objective sequence $\w$ in the context $\ctx$.
In \pref{def:uconcr}, we define $\uconcr(\w)$ as equal to $\concr(\w)$ if and only if $\concr(\w)$ contains scenarios starting from all states $s \subseteq \ctx$.
\pref{lem:uconcr-ctx} is used to over-approximate the initial context $\ctx$.

\begin{definition}[$\concr: \OS \to \powerset(\Sce)$]\label{def:concr}
Given $\w\in\OS$, $\concr(\w)$ is the set of minimal scenarios concretising $\w$ in the
context $\ctx$. It is defined as the largest set satisfying the following conditions:
\begin{enumerate}[(i)]
\item $\forall\delta\in\concr(\w), \modMF{\delta\in\Sce(\ctx)}$
\item $\forall\delta\in\concr(\w),
  \exists \phi:\indexes{\w}\to\indexes{\delta},
    (\forall n,m\in\indexes{\w}, n<m \Leftrightarrow \phi(n)\leq\phi(m)),$
\\\hspace*{2cm}$\forall n\in\indexes{\w}, \PHbounce(\w_n) \in \ctx\play\delta_{1..\phi(n)}$
\item $\forall\delta,\delta'\in\concr(\w)
				\card{\delta}\leq\card{\delta'} \Rightarrow
					\delta\neq\delta'_{1..\card{\delta}}$.
\end{enumerate}
\end{definition}

\noindent
The notation $\delta_{a..b}$ in the previous definition
denotes the subsequence of $\delta$ between indexes $a$ and $b$,
as defined on page \pageref{notations}.

\begin{definition}[$\uconcr: \OS \rightarrow \powerset(\Sce)$]
\label{def:uconcr}
  \[
  \modMF{\forall \w \in \OS,}
  \uconcr(\w) \DEF
  \begin{cases}
    \concr(\w) & \text{if } \forall s \in \PHl, s \subseteq \ctx, \exists \delta \in \concr(\w), \delta \in \Sce(s) \\
    \emptyset & \text{else.}
  \end{cases}
  \]
\end{definition}
% 
\begin{lemma}
\label{lem:uconcr-ctx}
  $\ctx \subseteq \ctx' \wedge \muconcr_{\ctx'}(\w) \neq \emptyset \Longrightarrow \muconcr_{\ctx}(\w) \neq \emptyset$.
\end{lemma}

For any objective $P$ and context $\ctx$, \pref{def:maxCont} gives the set of local states of automaton $\PHsort(P)$ that are required to solve $P$ in $\ctx$, denoted by $\gCont_\ctx(\PHsort(P), P)$.
\begin{definition}[$\gCont_\ctx : \Sigma \times \Obj \rightarrow \powerset(\PHproc)$]
  \label{def:maxCont}
  \begin{align*}
    \gCont_\ctx(a,P) \DEF
    \{ p \in \PHproc &\mid \exists ps \in \aBS(P), \exists b_i \in ps, b = a \wedge p = b_i \\
      & \vee b \neq a \wedge p \in \gCont_\ctx(a, \PHobj{b_j}{b_i}) \wedge b_j \in \PHget{\ctx}{b} \}
    \enspace.
  \end{align*}
\end{definition}

\modMF{
A graph of local causality $\myB = (\cwV, \cwE)$
is a graph where $\cwV \subseteq \Proc \cup \Obj \cup \sSol \cup \Sol$ is the set of vertices
and $\cwE \subseteq \cwV \times \cwV$ is the set of oriented edges.
Such a graph aims a describing the requirements of a given reachability
problem, that is, the reachability of $\w$ from the initial context $\ctx$.
Thus, a node in $\PHproc$ represents a local state required to play an action,
a node in $\Obj$ is an objective to reach a given local state,
a node in $\sSol$ is a set of hitters that have to be active simultaneously
to solve one step of a given objective,
and a node in $\Sol$ is a set of sets of hitters,
thus encompassing all requirement for the solving of a whole objective.
}

\modMF{
The edges of such a graph allow to draw links between these requirements.
Thus, an objective $P \in \Obj$ is solvable if
at least one related bounce sequence in $\BS(P)$ can be played;
therefore, we can consider all minimal sets of sets of hitters of these bounce sequences
(that is, the sets of $\aBS(P) \subseteq \Sol$, see \pref{def:aBS})
as requirements to solve this objective (\pref{eq:ESol1}).
Such a set can then naturally be split into as many sets of hitters (\pref{eq:ESol3})
which all have to be simultaneously active.
These sets of hitters can in turn be split into several local state nodes (\pref{eq:ESol2}).
This creates new requirements, as the reachability of a required local state $a_i$
is approximated by the ability to solve all objectives of the form
$\PHobjp{a}{j}{i} \in \Obj$ for all $a_j$ in the initial context (\pref{eq:EReq}).
Furthermore, we note that the solving of an objective $P$ may require
a local state of $\PHsort(P)$,
\ie $\gCont(\PHsort(P), P) \neq \emptyset$ (see \pref{def:maxCont});
in this case, $P$ is re-targeted (\pref{eq:ECont}).
Finally, \pref{eq:Vw} and \eqref{eq:Vproc}
force the graph to contain all the initial objectives
and the related local states,
and \pref{eq:VE} ensures the consistency of $\cwV$.
}

\modMF{
However, this graph $\myB$ alone is not sufficient to check the reachability of
an objective sequence $\w$ from an initial context $\ctx$;
indeed, the reachability of each intermediate required local state
is only checked starting from the context $\ctx$,
although this context may be modified by other objectives.
Therefore, such a reachability requires the building of the
\emph{saturated graph of local causality} (\pref{def:glc})
which is obtained by iteratively saturating the initial context with all
local states referenced in the graph,
until reaching a fixed point.
Such a fixed point always exists because the size of the context
is limited by the size of the initial PH model.
The set of local states contained in a graph of local causality
is given by:
\begin{align*}
  \allprocs(\cwV, \cwE) = (\cwV \cap \Proc) &\cup \{ \PHtarget(P) \mid P \in \cwV \cap \Obj \}\\
  & \cup \{ \PHbounce(P) \mid P \in \cwV \cap \Obj \wedge P \neq \w \}
\end{align*}
where, if the objective sequence $\w$ consists of only one element,
we can omit the hitter of this objective
provided that it does not appear elsewhere in the graph.
}

\begin{definition}
\label{def:glc}
  The \emph{\modMF{saturated} graph of local causality} $\cwB \DEF (\Bv, \Be)$ is defined as:
  $\cwB \DEF \lfp{\aB^\w_\ctx}{\myB}{\aB^\w_{\ctx \Cap \allprocs(\myB)}}$,
  where $\myB \DEF (\cwV, \cwE)$ is the smallest graph with
  $\cwV \subseteq \Proc \cup \Obj \cup \sSol \cup \Sol$ and $\cwE \subseteq \cwV \times \cwV$
  so that:
  \begin{align}
    \modMF{\toset{\w}} &\subseteq \cwV \label{eq:Vw} \\
    P \in \cwVObj &\Rightarrow \PHbounce(P) \in \cwV \label{eq:Vproc} \\
    (x, y) \in \cwE &\Rightarrow y \in \cwV \label{eq:VE} \\
    P \in \cwVObj \wedge pps \in \modMF{\aBS}(P) &\Rightarrow (P, pps) \in \cwE \label{eq:ESol1} \\
    \moda{pps \in \cwVSol \wedge A \in pps} &\moda{\Rightarrow (pps, A) \in \cwE} \label{eq:ESol3} \\
    A \in \cwVsSol \wedge a_i \in A &\Rightarrow (A, a_i) \in \cwE \label{eq:ESol2} \\
    a_i \in \cwVProc \wedge a_j \in \ctx &\Rightarrow (a_i, \PHobjp{a}{j}{i}) \in \cwE \label{eq:EReq} \\
%    a \in \cs \wedge a_i \in \cwVProc \wedge ps \in \csState(a_i) &\Rightarrow (a_i, ps) \in \cwE \label{eq:EPrio} \\
    P \in \cwVObj \wedge q \in \gCont_\ctx(\PHsort(P), P) &\Rightarrow (P, \PHobj{q}{\PHbounce(P)}) \in \cwE \label{eq:ECont} \!
  \end{align}
\end{definition}

\modMF{
In the saturated graph of local causality, a set of hitters $A \in \sSol$
is said to be \emph{coherent} (\pref{def:coherent})
if none of the local states in $A$ conflict with
a successor of $A$ in $\Bv \cap \Proc$,
that is, if there exists no local state node $a_j$ in the successors of $A$
and no element $a_i \in A$ so that $a_i \neq a_j$.
}
Then, \pref{th:approxinf} gives a sufficient condition for the concretisation
of an objective sequence in a given context,
which is derived immediately from the saturated graph of local causality,
and the condition that all sets of hitters are coherent.
A proof of this theorem is given in \pref{suppl:demoapproxinf}.
\begin{definition}[Coherent node]
\label{def:coherent}
  \modMF{
  In a given saturated graph of local causality $\cwB = (\Bv, \Be)$,
  a node $x \in \Bv$ is said to be \emph{coherent} if and only if:
  $x \in \Bv \cap \sSol \Rightarrow x$ has no successor $a_j \in \Bv \cap \Proc$
  so that $\exists a_i \in x$, $a_i \neq a_j$.
  }
\end{definition}

\begin{theorem}[Under-Approximation]
\label{th:approxinf}
  Given an AAN $(\PHs; \PHl; \PHa)$,
  a context $\ctx$ and an objective sequence $\w$,
  if the graph $\cwB$ contains no cycle,
  all objectives have at least one solution
  and all nodes are coherent,
  then $\uconcr(\w) \neq \emptyset$.
\end{theorem}

\modMF{
Regarding the complexity of the method,
computing the saturated graph of local causality is polynomial in the number of automata in $\PH$ and exponential in the number of local states in one automaton.
Checking the properties allowing to apply \pref{th:approxinf} is polynomial in the size of the graph.
Therefore, the building and checking process can be considered as polynomial in the size
of the AAN, provided that each automaton only contains a few local states.
We note that this is particularly true for biological models, where
each component usually contains a limited number of expression levels.
}

\modMF{
We note furthermore that in the case where
the method developed in this section is not conclusive,
it is possible to compute only a subset of $\Bv \cap \Sol$,
by removing some solutions from the initial graph of local causality $\myB$,
or by removing them from the saturated graph of local causality $\cwB$ and by trimming it.
In other words, this consists in ignoring some of the bounce sequences
for some of the objectives,
which intuitively cannot create false positives.
Indeed, \pref{th:approxinf} is then still valid on the partial graph obtained,
and this removal can lead to more conclusiveness by trimming parts of the graph
that were not necessary for the reachability
(especially cycles or unnecessary but inconclusive branches).
If one wants to try each possible subset of solutions,
then the overall method turns out to be exponential
in the number of solutions to each objective.
Our method can thus still be considered as efficient
compared to regular model-checking which is usually PSPACE-complete~\cite{Harel02}.
}

@@

>>>>

% vi:spell spelllang=en:
\section{Under-approximation of Reachability}\label{sec:sa}

\modLP{%
We present a static analysis that takes as input an AAN
$\PH = (\PHs; \PHl; \PHa)$,
an initial state, and a local state of an automata in $\PHs$, for identifying sufficient conditions
that ensure the existence of a scenario starting from the initial state and leading to a state
where the given local state is present.}

A classical approach for determining if a local state of an automaton can be reach from a
given initial state is to build sequences of transitions from the initial state until a state
containing the given local state is reached.
This is essentially what model-checkers do, and the complexity of such analysis (PSPACE-complete
\cite{Harel02}) makes it intractable on large systems, even with advanced symbolic approaches
\cite{PMR12-MSCS}.

Our approach relies on abstractions of scenarios that have been introduced in
\cite{PMR12-MSCS} for the static analysis of reachability in the Process Hitting framework,
a particular sub-class of AAN (see the Remark at the end of the previous section).
In this paper, we generalize the static analysis for the case of the under-approximation of
reachability in any AAN.

\todo{outline}

\subsection{Abstractions for Scenarios}
\label{ssec:abstr-sce}

The approach presented in this section is based on two complementary notions,
the \emph{objectives} and their \emph{local causality},
that are intertwined in so-called \emph{Local Causality Graphs}.

\subsubsection{Objectives}

An \emph{objective} (\pref{def:obj}) denotes the reachability of a local state (e.g., $a_j$) of a given automaton $a$
from the initial local state of that automaton (e.g., $a_i$).
Such an objective is written $\PHobjp{a}{i}{j}$.
Successive objectives are described with objective sequences (\pref{def:obj}).

\begin{definition}[Objective ($\Obj$) \& Objective Sequence ($\OS$)]
\label{def:obj}
  If $a \in \components$, the reachability of a local state $a_j$ from a local state $a_i$ is called an \emph{objective}, noted $\PHobj{a_i}{a_j}$.
  The set of all objectives is called $\Obj \DEF \{ \PHobj{a_i}{a_j} \mid a \in \components \wedge (a_i, a_j) \in \PHl_a \times \PHl_a \}$.
  For an objective $P = \PHobj{a_i}{a_j} \in \Obj$, we define: $\PHsort(P) \DEF
  a$, $\PHtarget(P)\DEF a_i$, $\PHbounce(P)\DEF a_j$,
  and $P$ is said \emph{trivial} if $a_i = a_j$.

@@

We define the partial ordering relation $\ltw$ between two objective sequences (\pref{def:ltw}) as
follows:
$w\ltw w'$ if and only if there exists a mapping between all the objective bounces of $w'$ in $w$
that preserve the sequentiality.

\todo{support compatibility}

\begin{definition}[$\ltw \subset \OS\times\OS$]\label{def:ltw}
$\w\ltw\w'$ if and only if
$|\w|\geq|\w'|$ and
there exists a mapping $\phi: \indexes{\w'}\mapsto \indexes{\w}$ such that
$\forall n,m \in\indexes{\w'}, n<m \Leftrightarrow \phi(n)<\phi(m)$,
and
$\forall n\in\indexes{\w'}, \bounce{\w'_n}=\bounce{\w_{\phi(n)}}$
\end{definition}

\begin{example}
\[\obj{b_0}{b_1}\concat\obj{a_0}{a_1}\concat\obj{b_1}{b_2}
\ltw
\obj{a_0}{a_1}\concat\obj{b_0}{b_2}
\ltw
\obj{b_0}{b_2}\]
\[\obj{b_0}{b_1}\not\ltw\obj{b_0}{b_2}\]
\end{example}

An objective sequence can be seen as an abstract representation of a set of scenarios that describe
(part of) the successive state changes of the automata.
We denote by $\concr(\w)$ (\pref{def:concr}) the set of scenarios concretising an objective sequence
$\w$ in the state $s\in\PHs$.
It is essentially all the scenarios for which there exists a mapping from the bounces of each
objective to the bounce of an action in the scenario which preserve the sequential ordering.

\todo{compatibility of $\w$ with $s$}

\begin{definition}[$\concr: \OS \to \powerset(\Sce)$]\label{def:concr}
Given $\w\in\OS$, $\concr(\w)$ is the set of scenarios concretising $\w$ in the
state $s$:
\begin{align*}
\concr(\w) \DEF \{ \delta\in\Sce(s)\mid & (\w^\vartriangle=\emptyseq \wedge
\delta=\emptyseq)  \vee (
\w^\vartriangle\neq \emptyseq\wedge
\\ &
 \wedge \exists \phi:\indexes{\w}\mapsto\indexes{\delta},
    (\forall n,m\in\indexes{\w}, n<m \Leftrightarrow \phi(n)\leq\phi(m))
\\ & \qquad
	\wedge \forall n\in\indexes{\w},
	  \bounce{\w_n} \in s\play\delta_{1..\phi(n)})
\}
\enspace,
\end{align*}
where $\omega^\vartriangle$ refers to the objective sequence $\omega$ where
trivial objectives have been removed.
The notation $\delta_{a..b}$ in the previous definition
denotes the subsequence of $\delta$ between indexes $a$ and $b$,
as defined on page \pageref{notations}.
\end{definition}

From \pref{def:ltw} and \pref{def:concr}, we derive that if 
$w\ltw w'$, then $\concr(w)\subseteq\concr(\w')$ (\pref{lem:ltw}).
\begin{lemma}\label{lem:ltw}
$\w\ltw\w' \Longrightarrow \concr(\w)\subseteq\concr(\w')$\enspace.
\end{lemma}

\subsubsection{Local Causality for Objectives}

The aim of searching for the existence of a scenario from a state $s$ leading to a state
where a given local state of an automaton $a_j$ is present
can be reformulated as the checking for the non-emptiness of $\concr(\obj{a_i}{a_j})$, where
$a_i=\get{s}{a}$.
A first hint for checking this emptiness is to look for actions that have to be played in order to
reach the state $i$ from $0$ within the automaton $a$.

Given an objective $P = \obj{a_i}{a_j} \in\Obj$, we define $\BS(P)$ the \emph{bounce sequences} of
$P$ as the set of minimal sequences of actions hitting $a$ in which the bounce of each action is
the target of the following action (\pref{def:bs}).

\begin{definition}[Bounce Sequence ($\BS$)]\label{def:bs}
A \emph{bounce sequence} $\zeta$  is a sequence of actions such that
$\forall n\in\indexes{\zeta}, n<|\zeta|,
\bounce{\zeta_{n}} = \target{\zeta_{n+1}}$.
$\BS$ denotes the set of bounce sequences.
We refer to the set of bounce sequences \emph{resolving} the objective $P$ as
$\BS(P)$:
\begin{align*}
\BS(\obj{a_i}{a_j}) \DEF \{ \zeta\in\BS\mid & \target{\zeta_1}=a_i\wedge
			    \bounce{\zeta_{|\zeta|}}=a_j \\
& \wedge \forall m,n\in\indexes{\zeta}, n>m, \bounce{\zeta_n}\neq\target{\zeta_m}
				\}\enspace.
\end{align*}
\end{definition}

Therefore,
$\BS(\obj{a_i}{a_i}) = \{\emptyseq\}$; and
$\BS(\obj{a_i}{a_j}) = \emptyset$ if there is no possibility to reach $a_j$ from
$a_i$.

\todo{example}

From \pref{def:bs}, we can derive that any scenario concretising an objective includes all the
actions of one of the bounce sequences of the objective (\pref{lem:bs-concr}).
\begin{lemma}
Given a state $s\in L$ and an objective $\obj{a_i}{a_j}$ with $\get{s}{a}=a_i$,
$\forall\delta\in\concr(\obj{a_i}{a_j})$,
$\exists\zeta\in\BS(\obj{a_i}{a_j})$ such that
$\exists \phi: \indexes{\zeta}\to\indexes{\delta}$
with
$\forall n,m\in\indexes{\zeta}, n<m \Leftrightarrow \phi(n)<\phi(m)$
and
$\forall n\in\indexes{\zeta}$, $\zeta_n = \delta_{\phi(n)}$.
\label{lem:bs-concr}
\end{lemma}

\todo{revise}
In the following, we denote: $\sSol = \powerset(\PHproc)$
\moda{%
and $\Sol = \powerset(\sSol)$.
}%
\modMF{%
The elements of $\sSol$ are sets of local states,
and will later represent sets of hitters of a given action
that are required to be active simultaneously in order to play this related action.
The elements in $\Sol$ are sets of sets of hitters,
and represent the requirements to play a whole bounce sequence.
}%

@@

\subsubsection{Local Causality Graph}

Given one initial objective, one could then recursively refine the objective into objective
sequences following its bounce sequences \cite{PMR12-MSCS}:
the refinement of objective $P$ with $\zeta\in\BS(P)$ would consists in prepending to $P$ the
objectives leading to the hitters of the actions in $\zeta$, in the same sequential order;
the generalisation of the refinement to an objective sequence would require to consider all possible
interleaving between the refinements.

These refinements can be summarized compactly with a so-called Local Causality Graph (LCG).
A LCG $\myB = (\cwV, \cwE)$
is a digraph where $\cwV \subseteq \Proc \cup \Obj \cup \sSol \cup \Sol$ is the set of vertices
and $\cwE \subseteq \cwV \times \cwV$ is the set of oriented edges.
\todo{explain nodes}
Such a graph aims a describing the requirements of a given reachability
problem, that is, the reachability of $\w$ from the initial context $\ctx$.
Thus, a node in $\PHproc$ represents a local state required to play an action,
a node in $\Obj$ is an objective to reach a given local state,
a node in $\sSol$ is a set of hitters that have to be active simultaneously
to solve one step of a given objective,
and a node in $\Sol$ is a set of sets of hitters,
thus encompassing all requirement for the solving of a whole objective.

\todo{explain edges}
The edges of such a graph allow to draw links between these requirements.
Thus, an objective $P \in \Obj$ is solvable if
at least one related bounce sequence in $\BS(P)$ can be played;
therefore, we can consider all minimal sets of sets of hitters of these bounce sequences
(that is, the sets of $\aBS(P) \subseteq \Sol$, see \pref{def:aBS})
as requirements to solve this objective (\pref{eq:ESol1}).
Such a set can then naturally be split into as many sets of hitters (\pref{eq:ESol3})
which all have to be simultaneously active.
These sets of hitters can in turn be split into several local state nodes (\pref{eq:ESol2}).
This creates new requirements, as the reachability of a required local state $a_i$
is approximated by the ability to solve all objectives of the form
$\PHobjp{a}{j}{i} \in \Obj$ for all $a_j$ in the initial context (\pref{eq:EReq}).
Furthermore, we note that the solving of an objective $P$ may require
a local state of $\PHsort(P)$,
\ie $\gCont(\PHsort(P), P) \neq \emptyset$ (see \pref{def:maxCont});
in this case, $P$ is re-targeted (\pref{eq:ECont}).
Finally, \pref{eq:Vw} and \eqref{eq:Vproc}
force the graph to contain all the initial objectives
and the related local states,
and \pref{eq:VE} ensures the consistency of $\cwV$.

\todo{motivate context}
A context (\pref{def:context}) extends the notion of state to a set of possible initial states:
\modMF{
to each automaton in the model, a context maps a set of local states in this automaton.
}
\begin{definition}[Context ($\Ctx$)]
\label{def:context}
  A \emph{context} $\ctx$ associates to each automaton in $\PHs$ a non-empty subset of its local states:
  $\forall a \in \PHs, \PHget{\ctx}{a} \subseteq \PHl_a \wedge \PHget{\ctx}{a} \neq \emptyset$.
  $\Ctx$ is the set of all contexts.
\end{definition}

Given a context $\ctx$ and an objective sequence $\w$, the corresponding LCG can be built
recursively starting from the objectives in $\w$ and recursively adding their possible refinements
following $\BS$ (\pref{def:glc}).

\begin{definition}
\label{def:glc}
Given a context $\ctx$ and a local state $p\in \Proc$,
the Local Causality Graph (LCG) for reachability under-approximation
$\cwB \DEF (\Bv, \Be)$
with
$\cwBNodes \subseteq \Proc \cup \Obj \cup \NSol \cup \powerset(\Proc)$
and
$\cwBEdges \subseteq \cwBNodes \times \cwBNodes$,
is the smallest graph such that:
\begin{enumerate}
\item
$p \in \cwBNodes$
\item
$a_i\in\cwBNodes\cap\Proc \Leftrightarrow \{ (a_i,\obj{a_j}{a_i}) \mid
a_j\in\ctx$ \\
\hspace*{3cm}
$\vee (a_j\neq a_i\wedge a_j\neq\w\wedge a_j\in\cwBNodes\cap\Proc)
\}\subseteq\cwBEdges$
\item
$P\in\cwBNodes\cap\Obj \Leftrightarrow 
	\{ (P,\langle P,nls\rangle) \mid nls\in\nsol(P) \}\subseteq\cwBEdges$
\item
$\langle P, nls \rangle\in\cwBNodes\cap\NSol \Leftrightarrow
	\{ (\langle P,nls\rangle, ls) \mid ls\in nls \}\subseteq\cwBEdges$
\item
$ls \in\cwBNodes\cap\powerset(\Proc) \Leftrightarrow
	\{ (ls, a_i) \mid a_i\in ls \}\subseteq\cwBEdges$
\item
$\obj{a_i}{a_j}\in \cwBNodes\cap\Obj \Rightarrow 
	\{(\obj{a_i}{a_j},\obj{a_k}{a_j}) \mid a_k\neq a_j,$
\\
\hspace*{4cm}
$a_k\in \conn_{(\cwBNodes,\cwBEdges)}(\langle\obj{a_i}{a_j},nls\rangle),$
\\
\hspace*{4cm}
$nls\in\nsol(\obj{a_i}{a_j}) \}\subset\cwBEdges$
\end{enumerate}

\todo{def conn, nsol}
\end{definition}

%
\begin{comment}
\noindent
For a given context $\ctx$, we note $a_i \in \ctx$ if and only if $a_i \in \PHget{\ctx}{a}$,
and for all $ps \in \powerset(\PHproc)$ or $ps \in \PHl$, $ps \subseteq \ctx \EQDEF \forall a_i \in ps, a_i \in \ctx$.
A sequence of actions $\delta$ is \emph{playable} in a context $\ctx$ if and only if 
$\exists s \subseteq \ctx, \delta \in \Sce(s)$.
We denote then: $\delta \in \Sce(\ctx)$,
and the play of $\delta$ in $\ctx$ is $\ctx \PHplay \delta = \ctx \Cap \ceil(\delta)$,
where $\ceil(\delta)$ is the set containing the last local state in the sequence $\delta$ (hitter or bounce) of every automaton mentioned in $\delta$.

\end{comment}

\subsection{Sufficient condition for reachability of a local state}
\label{ssec:ua}

\todo{explain rational}
\begin{itemize}
\item we want to identify an objective sequence that for sure has a concrete scenario
\item LCG is abstract representation of a set of objective sequences
\item we prove that if LCG has no cycle, and all its local states synchronizations are independent,
then such a sequence exists.
\item Note: as we are under-approximating, we de not require that this set is complete.
\end{itemize}

\todo{motivate $\uconcr$}
\begin{definition}[$\uconcr: \OS \rightarrow \powerset(\Sce)$]
\label{def:uconcr}
  \[
  \modMF{\forall \w \in \OS,}
  \uconcr(\w) \DEF
  \begin{cases}
    \concr(\w) & \text{if } \forall s \in \PHl, \concr(\w)\neq\emptyset\\
    \emptyset & \text{else.}
  \end{cases}
  \]
\end{definition}

\todo{revise}
\modMF{
In the saturated graph of local causality, a set of hitters $A \in \sSol$
is said to be \emph{coherent} (\pref{def:coherent})
if none of the local states in $A$ conflict with
a successor of $A$ in $\Bv \cap \Proc$,
that is, if there exists no local state node $a_j$ in the successors of $A$
and no element $a_i \in A$ so that $a_i \neq a_j$.
}
\begin{definition}[Independant synchronisations]
\label{def:coherent}
  In a LCG $\cwB = (\Bv, \Be)$,
  a node $x \in \Bv\cap\sSol$ is an \emph{independent synchronisation} if and only if
  $x$ has no successor $a_j \in \Bv \cap \Proc$ such that $\exists a_i \in x$, $a_i \neq a_j$.
\end{definition}

\todo{intuition on sufficient condition}
Then, \pref{th:approxinf} gives a sufficient condition for the reachability of 
a local state in a given context
of an objective sequence in a given context
which is derived immediately from the saturated graph of local causality,
and the condition that all sets of hitters are coherent.
A proof of this theorem is given in \pref{suppl:demoapproxinf}.

\begin{theorem}[Under-approximation]
\label{th:approxinf}
  Given an AAN $(\PHs; \PHl; \PHa)$,
  a context $\ctx$ and an objective sequence $\w$,
  if the LCG $\cwB$ contains no cycle,
  all objective nodes have at least one child,
  and all the synchronisations are independent,
  then $\uconcr(\w) \neq \emptyset$.
\end{theorem}

\modMF{
Regarding the complexity of the method,
computing the LCG is polynomial in the number of automata in $\PH$ and exponential in the number of local states in one automaton.
Checking the properties allowing to apply \pref{th:approxinf} is polynomial in the size of the graph.
Therefore, the building and checking process can be considered as polynomial in the size
of the AAN, provided that each automaton only contains a few local states.
We note that this is particularly true for biological models, where
each component usually contains a limited number of expression levels.
}

\modMF{
We note furthermore that in the case where
the method developed in this section is not conclusive,
it is possible to compute only a subset of $\Bv \cap \Sol$,
by removing some solutions from the initial graph of local causality $\myB$,
or by removing them from the saturated graph of local causality $\cwB$ and by trimming it.
In other words, this consists in ignoring some of the bounce sequences
for some of the objectives,
which intuitively cannot create false positives.
Indeed, \pref{th:approxinf} is then still valid on the partial graph obtained,
and this removal can lead to more conclusiveness by trimming parts of the graph
that were not necessary for the reachability
(especially cycles or unnecessary but inconclusive branches).
If one wants to try each possible subset of solutions,
then the overall method turns out to be exponential
in the number of solutions to each objective.
}

@@

\todo{integrate}
We also extend the override operator to contexts (\pref{def:ctxcap}).
\begin{definition}[$\Cap: \Ctx \times \powerset(\PHproc) \rightarrow \Ctx$]
\label{def:ctxcap}
  For any $\ctx\in\Ctx$ and set of local states $ps \in \powerset(\PHproc)$,
  the override of $\ctx$ by $ps$ is noted $\ctx \Cap ps$ and is defined by:
  \[ \forall a \in \PHs, \PHget{(\ctx \Cap ps)}{a} \DEF
  \begin{cases}
    \{ p \in ps \mid \PHsort(p)=a \} & \text{if } \exists p \in ps, \PHsort(p)=a,\\
    \PHget{\ctx}{a} & \text{else.}
  \end{cases}
  \]
\end{definition}
